---
title: "On the usage of the  pbkrtest package"
author: "Søren Højsgaard^1^ and Ulrich Halekoh^2^"
date: "`r Sys.Date()`"
geometry: margin=3cm
output:
  html_document:
    fig_caption: yes
    number_sections: yes
    toc: yes
  pdf_document:
    dev: cairo_pdf
    fig_caption: yes
    number_sections: yes
    toc: yes  
documentclass: article
header-includes:
- \usepackage{fancyhdr}
- \usepackage{wrapfig}
- \usepackage{float}
- \pagestyle{fancy}
- \fancyfoot[C]{\thepage}
- \usepackage{hyperref}
- \hypersetup{colorlinks=true}
- \hypersetup{linktoc=all}
- \hypersetup{linkcolor=blue}
- \usepackage{pdflscape}
- \usepackage{booktabs}
- \newcommand{\blandscape}{\begin{landscape}}
- \newcommand{\elandscape}{\end{landscape}}
- \renewcommand\thesection{\arabic{section}}
link-citations: yes
csl: frontiers.csl
resource_files:
- vignettes/rbase.png
- vignettes/rstudio.png
- vignettes/rstudio panes.png
vignette: |
  %\VignetteIndexEntry{Intro_pbkrtest}
  %\usepackage[utf8]{inputenc}
  %\VignetteEngine{knitr::rmarkdown_notangle}
  %\VignetteEncoding{UTF-8}
---

```{r, echo=FALSE}
out_type <- knitr::opts_knit$get("rmarkdown.pandoc.to")

r = getOption("repos")
r["CRAN"] = "https://cran.rstudio.com/"
#r["CRAN"] = "https://cloud.r-project.org/"
#r["CRAN"] = "https://ftp.iitm.ac.in/cran/"
options(repos = r)
```

```{r, results='asis', echo=FALSE}
switch(out_type,
    html = {cat("
  <p> 1. Head of Department of Mathematical Sciences and Associate Professor of statistics at Aalborg University  (sorenh@math.aau.dk).</p>
  
  <p> 2. Associate Professor, Department of Public Health, University of Southern Denmark, Denmark (uhalekoh@health.sdu.dk).</p>  
                ")},
    latex = cat("\\begin{center}
1. Head of Department of Mathematical Sciences and Associate Professor of statistics at Aalborg University  (sorenh@math.aau.dk).

2. Associate Professor, Department of Public Health, University of Southern Denmark, Denmark (uhalekoh@health.sdu.dk)

\\end{center}" )
)
```

\begin{center}
\vspace{6pt}
\hrule
\end{center}


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      comment = "",
                      fig.cap = "")
```

```{r echo = FALSE, message=FALSE}
library(pbkrtest)
```

# Introduction

The \code{shoes} data is a list of two vectors, giving the wear of
shoes of materials A and B for one foot each of ten boys.

```{r}
data(shoes, package="MASS")
shoes
```


A plot clearly reveals that boys wear their shoes differently.

```{r}
plot(A~1, data=shoes, col="red",lwd=2, pch=1, ylab="wear", xlab="boy")
points(B~1, data=shoes, col="blue", lwd=2, pch=2)
points(I((A+B)/2)~1, data=shoes, pch="-", lwd=2)
```



One option for testing the effect of materials is to make a paired
$t$--test. The following forms are equivalent:

```{r}
r1<-t.test(shoes$A, shoes$B, paired=T)
r2<-t.test(shoes$A-shoes$B)
r1
```


To work with data in a mixed model setting we create a dataframe, and
for later use we also create an imbalanced version of data:

```{r}
boy <- rep(1:10,2)
boyf<- factor(letters[boy])
mat <- factor(c(rep("A", 10), rep("B",10)))
## Balanced data:
shoe.b <- data.frame(wear=unlist(shoes), boy=boy, boyf=boyf, mat=mat)
head(shoe.b)
## Imbalanced data; delete (boy=1, mat=1) and (boy=2, mat=b)
shoe.i <-  shoe.b[-c(1,12),]
```


We fit models to the two datasets:

```{r}
lmm1.b  <- lmer( wear ~ mat + (1|boyf), data=shoe.b )
lmm0.b  <- update( lmm1.b, .~. - mat)
lmm1.i  <- lmer( wear ~ mat + (1|boyf), data=shoe.i )
lmm0.i  <- update(lmm1.i, .~. - mat)
```


The asymptotic likelihood ratio test shows stronger significance than
the $t$--test:

```{r}
anova( lmm1.b, lmm0.b, test="Chisq" ) ## Balanced data
anova( lmm1.i, lmm0.i, test="Chisq" ) ## Imbalanced data
```

# Kenward--Roger approach


The Kenward--Roger approximation is exact for the balanced data in the
sense that it produces the same result as the paired $t$--test.

```{r}
( kr.b<-KRmodcomp(lmm1.b, lmm0.b) )
```

```{r}
summary(kr.b)
```


Relevant information can be retrieved with

```{r}
getKR(kr.b, "ddf")
```


For the imbalanced data we get

```{r}
(kr.i<-KRmodcomp(lmm1.i, lmm0.i))
```


Notice that this result is similar to but not identical to the paired
$t$--test when the two relevant boys are removed:

```{r}
shoes2 <- list(A=shoes$A[-(1:2)], B=shoes$B[-(1:2)])
t.test(shoes2$A, shoes2$B, paired=T)
```



# Parametric bootstrap

Parametric bootstrap provides an alternative but many simulations are
often needed to provide credible results (also many more than shown
here; in this connection it can be useful to exploit that computings
can be made en parallel, see the documentation):

```{r}
(pb.b <- PBmodcomp(lmm1.b, lmm0.b, nsim=500, cl=2) )
```

```{r}
summary( pb.b )
```



For the imbalanced data, the result is similar to the result from the
paired $t$ test.

```{r}
(pb.i<-PBmodcomp(lmm1.i, lmm0.i, nsim=500, cl=2))
```

```{r}
summary(pb.i)
```


\appendix

# Matrices for random effects

The matrices involved in the random effects can be obtained with

```{r}
shoe3 <- subset(shoe.b, boy<=5)
shoe3 <- shoe3[order(shoe3$boy), ]
lmm1  <- lmer( wear ~ mat + (1|boyf), data=shoe3 )
str( SG <- get_SigmaG( lmm1 ), max=2)
```



```{r}
round( SG$Sigma*10 )
```


```{r}
SG$G
```
